Pattern matching for\\ $(213,231)$-avoiding permutations

### 2015-02-11 article.tex
modification : add lemma for section 4

%first
\DeclareMathOperator{\firsta}{L}
\newcommand{\first}[1]{\firsta({#1})}

\DeclareMathOperator{\factora}{F}
\newcommand{\factor}[1]{\factora({#1})}

@@

@@

This section focuses on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.
We need to consider a specific decomposition $F$ of $\sigma$ into factors.
For $\sigma \in S_k$, the factorization
$\sigma = F(p) F(p-1) \ldots F(1)$ is defined as follows:
(i) $\sigma[k]$ is the last letter of $F(1)$,
(ii) for every $1 \leq i < k$,
if $s_\sigma[i] = s_\sigma[i+1]$ then
$\sigma[i]\sigma[i+1]$ is part of the same factor.
For every factor $F(j)$ of the decomposition, we let
$L(j)$ stand for the leftmost letter of $F(j)$.
For example,
for $\sigma = 981237654$,
we have
$s_\sigma = ddaaaddd$ and hence
$\sigma = F(3) F(2) F(1)$ with
$F(1) =7654$, $F(2) = 123$ and $F(3) = 98$.
Furthermore,
$L(1) = 7$, $L(2) = 1$ and $L(3) = 9$.

To ease the lemma and its proof we define the set $\SET{\pmotif}{\ptext}{i}{j}$ such that if $s \in S$, then $s$ is an occurrence of $\factor{i} \factor{i-1} \cdots \factor{1}$ in $\ptext[j:]$ and $s[\first{i}]=\ptext[j]$.

\begin{lemma}
Given a permutation $\pmotif$ and $\factor{i}$ an ascent (descent) factor.
Let's choose a subsequence $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that $s[\first{i-1}]$ is minimal (maximal), given a subsequence
$s' \in \SET{\pmotif}{\ptext}{i}{j}$ and a subsequence $t=t's'$ of $\ptext$  if $t$ is an occurrence of $\factor{i+1} \factor{i} \factor{i-1} \cdots \factor{1}$ then the subsequence $t's$ is an occurrence of $\factor{i+1} \factor{i} \factor{i-1} \cdots \factor{1}$ in $\ptext$.
\end{lemma}

\begin{proof}
Given a permutation $\pmotif$ and $\factor{i}$ an ascent (descent) factor.
Let's choose a subsequence $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that $s[\first{i-1}]$ is minimal (maximal).
Let's $s' \in \SET{\pmotif}{\ptext}{i}{j}$ and a subsequence $t=t's'$ of $\ptext$ such that $t$ is an occurrence of $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$. 
Remark that every element of $t'$ must be superior (inferior) to every element of $s'$, and that the maximal element of $s'$ is the element matched to $t[\first{i-1}]$. Thus if we have a subsequence $s \in \SET{\pmotif}{\ptext}{i}{j}$ such that $s[\first{i-1}]$ is minimal (maximal), then $s[\first{i-1}]\leq s'[\first{i-1}]$, so every element of $'t$ is also superior (inferior) to every element of $s$. Thus the subsequence $t's$ is an occurrence of $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$ in $\ptext$.
\end{proof}

\begin{corollary}
Let's choose a subsequence $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that $s[\first{i-1}]$ is minimal (maximal). There is a match of $\pmotif$ in $\ptext$ with element $\first{i}$ matched to $\ptext[j]$ if and only if there exists $t$ such that $ts$ is a match of $\pmotif$ in $\ptext$. 
\end{corollary}

This corollary give us a safe way to choose a match.

@@

Consider the following problem :\\
$LM(i,j)$ = the safest match of $\first{i-1}$
	if there exists a match of $\pmotif[\first{i}:]$
	with $\ptext[j:]$
	and element $\first{i}$
	is matched with element at index $j$.\\

@@

\textbf{BASE :} \\
$
LM(1,j)=
\begin{cases}
		MIN_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } &
			\text{if $\stripe{\pmotif}{\first{1}}=\ustep$} \\
		\text{ $|\factor{1}| \leq LIS(j,j',\ptext[j'])  $}\}& \\
		&\\
		MAX_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } &
			\text{if $\stripe{\pmotif}{\first{1}}=\dstep$} \\
		\text{ $|\factor{1}| \leq LDS(j,j',\ptext[j']) $}\}&\\
\end{cases}
$

\textbf{STEP :} \\
$
LM(i,j)=
\begin{cases}
	MIN \{0\} \cup  I(i,j) &
	\text{if $\stripe{\pmotif}{\first{i}}=\ustep$}\\
	MAX \{0\} \cup  D(i,j) &
	\text{if $\stripe{\pmotif}{\first{i}}=\dstep$}\\
\end{cases}
$

With $I(i,j)$ ($D(i,j)$) the set of element such that
if  $j' \in I(i,j)$ then there exists
a match of $\pmotif[\first{i-1}:]$ in $\ptext[j':]$ and
the element $\first{i-1}$ is matched with element at index $j'$ and
there is a match of $\factor{i}$ in $\ptext[j:j'-1]$ and
the element $\first{i}$ is matched with element at index $j$ and every element is inferior (superior) to the match of
$\pmotif[\first{i-1}:]$ in $\ptext[j':]$.
Formally we define $I(s,i)$ and $D(s,i)$ such as:\\

$I(i,j)=\{j'|\text{$j<j'$ and $LM(i-1,j')\neq 0$}$ \\
\indent \indent \indent and  $|\factor{i}| \leq LIS(j,j'-1,\first{i-1})\}$ \\

$D(i,j)=\{j'|\text{$j<j'$ and $LM(i-1,j')\neq 0$}$ \\
\indent \indent \indent and  $|\factor{i}| \leq LDS(j,j'-1,\first{i-1}) \}$ \\

With LIS(i,j,k) (LDS(i,j,k)) is the longest increasing (decreasing) sequence in $\ptext$ starting at $i$ and ending at $j$,
with every element of this sequence
inferior (superior) to $k$.
LIS and LDS can be computed in $O(|\ptext|^2*\log(\log(|\ptext|)))$ (see \cite{Bespamyatnikh00enumeratinglongest}).\\

In the base case,
we are looking for a match for the first factor.
Each factor is either ascent or descent.
If the factor is an ascent (descent)
we have to find an increasing (decreasing) subsequence
in the text of same size or longest that
the size of the factor,
and to find the optimal solution
we must assure that the last element
of the sequence is minimal (maximal).\\
For the induction, it is the same idea
except that we must assure that
every element of the subsequence is superior (inferior)
to the maximum (minimal) element of the rest of the match
ie inferior (superior) to the minimal (maximal) element of the rest of the match
which is given by $LM$ of
the previous factor.\\

There exists a match of $\pmotif$ in $\ptext$ if and only if
there exists a $LM(n,i)\neq 0$ for $1 \leq i \leq |\ptext|$,
with $n$ the number of factor in $\pmotif$.
Moreover the basic case can be computed in $O(|\ptext|^2)$
and the induction on $O(|\pmotif||\ptext|^2)$.
\qed
\end{proof}

>>>>

%first
\DeclareMathOperator{\firsta}{L}
\newcommand{\first}[2]{\firsta_{{#1}}({#2})}

\DeclareMathOperator{\firstia}{LI}
\newcommand{\firsti}[2]{\firstia_{{#1}}({#2})}

\DeclareMathOperator{\factora}{F}
\newcommand{\factor}[2]{\factora_{{#1}}({#2})}

@@

%LM
\DeclareMathOperator{\LMa}{LM}
\newcommand{\LM}[4]{\LMa_{{#1}}^{{#2}}(#3,#4)}

\DeclareMathOperator{\AFa}{AF}
\newcommand{\AF}[4]{\AFa_{{#1}}^{{#2}}(#3,#4)}

\DeclareMathOperator{\DFa}{DF}
\newcommand{\DF}[4]{\DFa_{{#1}}^{{#2}}(#3,#4)}

@@

This section focuses on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.
We need to consider a specific decomposition $\factora$ of $\sigma$ into factors.
For $\sigma \in S_k$, the factorization
$\sigma = \factor{\pmotif}{p} \factor{\pmotif}{p-1} \ldots \factor{\pmotif}{1}$ is defined as follows:
(i) $\sigma[k]$ is the last letter of $\factor{\pmotif}{1}$,
(ii) for every $1 \leq i < k$,
if $s_\sigma[i] = s_\sigma[i+1]$ then
$\sigma[i]\sigma[i+1]$ is part of the same factor.
For every factor $\factor{\pmotif}{j}$ of the decomposition, we let
$\first{\pmotif}{j}$ stand for the leftmost letter of $\factor{\pmotif}{j}$,
and $\firsti{\pmotif}{j}$ stand for the index of the leftmost letter of $\factor{\pmotif}{j}$.
For example,
for $\sigma = 981237654$,
we have
$s_\sigma = ddaaaddd$ and hence
$\sigma = \factor{\pmotif}{3} \factor{\pmotif}{2} \factor{\pmotif}{1}$ with
$\factor{\pmotif}{1} =7654$, $\factor{\pmotif}{2} = 123$ and $\factor{\pmotif}{3} = 98$.
Furthermore,
$\first{\pmotif}{1} = 7$, $\first{\pmotif}{2} = 1$, $\first{\pmotif}{3} = 9$,
$\firsti{\pmotif}{1} = 6$, $\firsti{\pmotif}{2} = 3$ and $\firsti{\pmotif}{3} = 1$.

\begin{remark}
A Factor is either an increasing or a decreasing subsequence.
\end{remark}

\begin{lemma}
\label{lemma:whereIsMax}
Given a permutation $\sigma \in \AV(213,231)$ and a suffix of its decomposition $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$, if
$\stripe{\sigma}{\firsti{\sigma}{i}}=\ustep$ ($\stripe{\sigma}{\firsti{\sigma}{i}}=\dstep$)  the maximal (minimal) element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ is $\first{\sigma}{i-1}$.
\end{lemma}

\begin{proof}
Take the right most element of $\factor{\sigma}{i}$, this element is an ascent (descent),
so it is inferior (superior) to the next element which is $\first{\sigma}{i-1}$. Because $\factor{\sigma}{i}$ is an increasing (decreasing) subsequence, every element of $\factor{\sigma}{i}$ is inferior (superior) to $\first{\sigma}{i-1}$.
By lemma \ref{lemma:first element is 1 or n}, $\first{\sigma}{i-1}$ is the maximal (minimal) element of $\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ thus $\first{\sigma}{i-1}$ is the maximal (minimal) element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$.

\end{proof}

To ease the next lemma and its proof we define the set $\SET{\pmotif}{\ptext}{i}{j}$ such that if $s \in S$, then $s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$ and the element matching $\first{\pmotif}{i}$ is $\ptext[j]$.

\begin{lemma}
Given a permutation $\pmotif$ and $\factor{\pmotif}{i}$ an ascent (descent) factor.
Let's $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that the element matching $\first{\pmotif}{i-1}$ is minimal (maximal), for all subsequence 
$s' \in \SET{\pmotif}{\ptext}{i}{j}$ and for all subsequence of $\ptext$ $t$, such that $t=t's'$, if $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ then the subsequence $t's$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$.
\end{lemma}

\begin{proof}
Given a permutation $\pmotif$ and $\factor{\pmotif}{i}$ an ascent (descent) factor.
Let's $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that the element matching $\first{\pmotif}{i-1}$ is minimal (maximal).
Let's $s' \in \SET{\pmotif}{\ptext}{i}{j}$ and $t$ subsequence of $\ptext$, such that $t=t's'$ and $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ in $\pi$. 
Remark that every element of $t'$ is superior (inferior) to every element of $s'$.
Note that the element matching $\first{\pmotif}{i-1}$ in $s$ is minimal (maximal) and it is the maximal (minimal) element of the subsequence (\ref{lemma:whereIsMax}), thus
it is inferior (superior) to the element matching $\first{\pmotif}{i-1}$ in $s'$.
So every element of $t'$ is also superior (inferior) to every element of $s$. Thus the subsequence $t's$ is an occurrence of $\pmotif[\first{\pmotif}{i+1}:]$ in $\ptext$.
\end{proof}

\begin{corollary}
Let's choose a subsequence $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that $s[\first{\pmotif}{i-1}]$ is minimal (maximal). There is a match of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ matched to $\ptext[j]$ if and only if there exists $t$ such that $ts$ is a match of $\pmotif$ in $\ptext$. 
\end{corollary}

This corollary gives one a safe way to choose a match. Indeed if one does not find a match such that a well chosen subsequence is a suffix then there is no match at all.

@@

Consider the following problem :\\
$
\LM{\pmotif}{\ptext}{i}{j} = 
\begin{cases}
	\text{The maximal value of a subsequence} 
		& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\ustep$ } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & \\	
	\text{ and that minimize the maximal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\

	\text{The minimal value of a subsequence} 
		& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\dstep$ } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & \\	
	\text{ and that maximize the minimal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\
\end{cases}
$

@@

\textbf{BASE :} \\
$
\LM{\pmotif}{\ptext}{1}{j} = 
\begin{cases}
		\min_{j<j'}\{\infty\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\ustep$ }  \\
		\text{ $|\factor{\pmotif}{1}| \leq LIS(j,j',\ptext[j'])  $}\}& \\
		&\\
		\max_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\dstep$ } \\
		\text{ $|\factor{\pmotif}{1}| \leq LDS(j,j',\ptext[j']) $}\}&\\
\end{cases}
$

\textbf{STEP :} \\
$
\LM{\pmotif}{\ptext}{i}{j} =
\begin{cases}
	\min \{\infty\} \cup  \AF{\pmotif}{\ptext}{i}{j} &
	\text{if $\stripe{\pmotif}{\first{\pmotif}{i}}=\ustep$}\\
	\max \{0\} \cup  \DF{\pmotif}{\ptext}{i}{j} &
	\text{if $\stripe{\pmotif}{\first{\pmotif}{i}}=\dstep$}\\
\end{cases}
$

With $\AF{\pmotif}{\ptext}{i}{j}$ ($\DF{\pmotif}{\ptext}{i}{j}$) the set of element such that
if  $\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$ then there exists
a match of $\pmotif[\firsti{\pmotif}{i-1}:]$ in $\ptext[k:]$ and
the element $\first{\pmotif}{i-1}$ is matched with element $\ptext[k]$ and
there is a match of $\factor{\pmotif}{i}$ in $\ptext[j:k-1]$ and
the element $\first{\pmotif}{i}$ is matched with element $\ptext[j]$ and every element is inferior (superior) to the match of
$\pmotif[\firsti{\pmotif}{i-1}:]$ in $\ptext[k:]$.
Formally we define $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ such as:\\

$\AF{\pmotif}{\ptext}{i}{j}=\{j'|\text{$j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq 0$}$ \\
\indent \indent \indent and  $|\factor{\pmotif}{i}| \leq LIS_{\ptext}(j,j'-1,\first{\pmotif}{i-1})\}$ \\

$\DF{\pmotif}{\ptext}{i}{j}=\{j'|\text{$j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq \infty$}$ \\
\indent \indent \indent and  $|\factor{\pmotif}{i}| \leq LDS_{\ptext}(j,j'-1,\first{\pmotif}{i-1}) \}$ \\

With $LIS_{\ptext}(i,j,k)$ ($LDS_{\ptext}(i,j,k)$) is the longest increasing (decreasing) sequence in $\ptext$ starting at $i$ and ending at $j$,
with every element of this sequence
inferior (superior) to $k$.
$LIS_{\ptext}$ and $LDS_{\ptext}$ can be computed in $O(|\ptext|^2*\log(\log(|\ptext|)))$ (see \cite{Bespamyatnikh00enumeratinglongest}).\\

In the base case,
one is looking for a match for the first factor.
If the factor is an ascent (descent),
one has to find an increasing (decreasing) subsequence
in the text of same size or longest that
the size of the factor,
and to find the best solution
one must assure that the last element
of the sequence is minimal (maximal).\\
For the induction, it is the same idea
except that one must assure that
every element of the subsequence is superior (inferior)
to the maximum (minimal) element of the rest of the match which is given by $LM$ of
the previous factor.\\

There exists a match of $\pmotif$ in $\ptext$ if and only if
there exists a $LM(n,i)\neq 0$ and $LM(n,i)\neq \infty$ for $1 \leq i \leq |\ptext|$,
with $n$ the number of factor in $\pmotif$.
Moreover the basic case can be computed in $O(n^2)$
and the induction on $O(kn^2)$ time and space.
\qed
\end{proof}
